---
title: "Lab 11 Homework"
author: "Jodie Cheun"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

**In this homework, you should make use of the aesthetics you have learned. It's OK to be flashy!**

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
``` 

```{r}
options(scipen = 999)
```

## Resources
The idea for this assignment came from [Rebecca Barter's](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/) ggplot tutorial so if you get stuck this is a good place to have a look.  

## Gapminder
For this assignment, we are going to use the dataset [gapminder](https://cran.r-project.org/web/packages/gapminder/index.html). Gapminder includes information about economics, population, and life expectancy from countries all over the world. You will need to install it before use. This is the same data that we will use for midterm 2 so this is good practice.
```{r}
#install.packages("gapminder")
library("gapminder")
```
## Questions
The questions below are open-ended and have many possible solutions. Your approach should, where appropriate, include numerical summaries and visuals. Be creative; assume you are building an analysis that you would ultimately present to an audience of stakeholders. Feel free to try out different `geoms` if they more clearly present your results.  

**1. Use the function(s) of your choice to get an idea of the overall structure of the data frame, including its dimensions, column names, variable classes, etc. As part of this, determine how NAs are treated in the data.**  
```{r}
gapminder <- gapminder %>% clean_names()
gapminder
```

```{r}
dim(gapminder)
```

```{r}
names(gapminder)
```
```{r}
summary(gapminder)
```

```{r}
glimpse(gapminder)
```
As shown we have a combination of factors, numeric, and categorical data. 
```{r}
sum(is.na(gapminder))
```

```{r}
miss_var_summary(gapminder)
```
As shown this data does not contain any NA values, nor any placeholder values that should be counted as NAs. The data is neat and follows tidyverse format. 

**2. Among the interesting variables in gapminder is life expectancy. How has global life expectancy changed between 1952 and 2007?**

```{r}
gapminder %>% 
  mutate(year=as.factor(year)) %>% 
  select(country,year,life_exp) %>% 
  filter(year=="1952"|year=="2007") %>% 
  pivot_wider(names_from ="year",values_from="year")
```
As shown above, we see an average trend of increasing `life_exp` from 1952 to 2007 per each of the analyzed countries. 

**3. How do the distributions of life expectancy compare for the years 1952 and 2007?**
```{r}
gapminder %>% 
  mutate(year=as.factor(year)) %>% 
  filter(between(year,"1952","2007"))%>%
  ggplot(aes(x=year,y=life_exp))+
  geom_col(na.rm=T)+
  labs(title= "Range of Life Expectancy",
       x="Year",
       y="Life Expectancy")
```
As shown, we observe a steady increase in life expentancy from 1952 to 2007. 
Note to self: remember that by changing the years to factors instead of numerical data, we must put "" when selecting years we want to filter or specifically analyze. 

**4. Your answer above doesn't tell the whole story since life expectancy varies by region. Make a summary that shows the min, mean, and max life expectancy by continent for all years represented in the data.**
Min, mean, and max can be represented by a boxplot. Note, `summary` here indicates we are looking for a numerical summary 

```{r}
gapminder %>% 
  mutate(year=as.factor(year)) %>% 
  group_by(continent,year) %>% 
  summarise(min_life_exp=min(life_exp),
            mean_life_exp=mean(life_exp),
            max_life_exp=max(life_exp))
```
**5. How has life expectancy changed between 1952-2007 for each continent?**
```{r}
gapminder %>% 
  mutate(year=as.factor(year)) %>% 
  ggplot(aes(x=continent,y=life_exp,fill=continent))+
  geom_boxplot(na.rm=T,alpha=0.5)+
  labs(title= "Range of Life Expectancy",
       x="Continent",
       y="Life Expectancy")
```
Confusion, if we do not include a certain year to grab the data from, does this mean these box plots are the min,mean,and max from all the data from all the years measured?

**6. We are interested in the relationship between per capita GDP and life expectancy; i.e. does having more money help you live longer?**
```{r}
names(gapminder)
```


```{r}
gapminder %>% 
   mutate(year=as.factor(year)) %>% 
  ggplot(aes(y=gdp_percap,x=life_exp,color=continent))+
  geom_jitter(size=0.65,alpha=0.8)+
  labs(title= "Does Money Make You Live Longer?",
       x="Life Expectancy",
       y="Income")
```
There is a subtle trend that shows longer life expectancy can be associated with a higher income. However, additional data is needed to make a more confident conclusion, especially due to the outlier points of younger individuals with higher incomes having shorter life expectancies. 

**7. Which countries have had the largest population growth since 1952?**
Numerical Summary 
```{r}
gapminder %>% 
  mutate(year=as.factor(year)) %>% 
  select(country,year,pop) %>% 
  filter(year== "1952" | year =="2007") %>% 
  pivot_wider(names_from =year,names_prefix = "yr_", values_from=pop) %>% 
  mutate(pop_growth=yr_2007-yr_1952) %>% 
  arrange(desc(pop_growth))
```
As shown in the numerical summary above, the top 3 countries with the highest population growth between 1952 and 2007 is China, India, and the United States. 

Plot
```{r}
gapminder %>% 
  mutate(year=as.factor(year)) %>% 
  select(country,year,pop) %>% 
  filter(year== "1952" | year =="2007") %>% 
  filter(country=="China"|country=="India"|country=="United States") %>% 
  pivot_wider(names_from =year,names_prefix = "yr_", values_from=pop) %>% 
  mutate(pop_growth=yr_2007-yr_1952) %>% 
  ggplot(aes(x=country,y=pop_growth))+
  geom_col(na.rm=T, color="black", fill="light blue")+
  labs(title="Population Growth from 1952 to 2007",
       x="Countries",
       y= "Population Growth")
```

**8. Use your results from the question above to plot population growth for the top five countries since 1952.**

```{r}
gapminder %>% 
  mutate(year=as.factor(year)) %>% 
  select(country,year,pop) %>% 
  filter(year== "1952" | year =="2007") %>% 
  pivot_wider(names_from =year,names_prefix = "yr_", values_from=pop) %>% 
  mutate(pop_growth=yr_2007-yr_1952) %>% 
  arrange(desc(pop_growth))
```
The top 5 countries with the highest population growth between 1952 to 2007 (in decreasing order) is China, India, United States, Indonesia, and Brazil. 

```{r}
gapminder %>% 
  mutate(year=as.factor(year)) %>% 
  select(country,year,pop) %>% 
  filter(year== "1952" | year =="2007") %>% 
  filter(country=="China"|country=="India"|country=="United States"|country=="Indonesia"|country=="Brazil") %>% 
  pivot_wider(names_from =year,names_prefix = "yr_", values_from=pop) %>% 
  mutate(pop_growth=yr_2007-yr_1952) %>% 
  ggplot(aes(x=country,y=pop_growth))+
  geom_col(na.rm=T, color="black", fill="pink")+
  labs(title="Population Growth from 1952 to 2007",
       x="Countries",
       y= "Population Growth")
```

**9. How does per capita GDP growth compare between these same five countries?**
```{r}
gapminder %>% 
  mutate(year=as.factor(year)) %>% 
  select(country,year,gdp_percap) %>% 
  filter(year== "1952" | year =="2007") %>% 
  filter(country=="China"|country=="India"|country=="United States"|country=="Indonesia"|country=="Brazil") %>% 
  pivot_wider(names_from =year,names_prefix = "yr_", values_from=gdp_percap) %>% 
  mutate(gdp_growth=yr_2007-yr_1952) %>% 
  ggplot(aes(x=country,y=gdp_growth))+
  geom_col(na.rm=T, color="black", fill="pink")+
  labs(title="GDP Growth from 1952 to 2007",
       x="Countries",
       y= "GDP Growth")
```

**10. Make one plot of your choice that uses faceting!**    
```{r}
str(gapminder)
```


```{r}
gapminder %>% 
  mutate(year=as.factor(year)) %>% 
  ggplot(aes(x=continent,y=life_exp,fill=continent))+
  geom_boxplot(na.rm=T,alpha=0.5)+
  labs(title= "Range of Life Expectancy",
       x="Continent",
       y="Life Expectancy")+
  facet_grid(.~life_exp)
```



## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 