---
title: "Homework 10"
author: "Jodie Cheun"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(naniar)
``` 

## Desert Ecology
For this assignment, we are going to use a modified data set on [desert ecology](http://esapubs.org/archive/ecol/E090/118/). The data are from: S. K. Morgan Ernest, Thomas J. Valone, and James H. Brown. 2009. Long-term monitoring and experimental manipulation of a Chihuahuan Desert ecosystem near Portal, Arizona, USA. Ecology 90:1708.
```{r}
deserts <- read_csv("data/surveys_complete.csv") %>% clean_names()
```

1. Use the function(s) of your choice to get an idea of its structure, including how NA's are treated. Are the data tidy? 
```{r}
names(deserts)
```

```{r}
summary(deserts)
```
```{r}
miss_var_summary(deserts)
```
```{r}
deserts
```
Yes the data does look tidy. Each of the variables does not contain additional information. And each observation is unique in its own row/cell. 

2. How many genera and species are represented in the data? What are the total number of observations? Which species is most/ least frequently sampled in the study?
```{r}
names(deserts)
```

```{r}
deserts %>% 
  n_distinct("genus")
```

```{r}
deserts %>% 
  n_distinct("species")
```

In the data, we have a total of 34786 distinct observations for the `genus` and `species` variables. 

```{r}
deserts %>% 
  count(species) %>% 
  arrange(desc(n))
```
The most studied species, based on the number of recorded observations is the merriami species. The least studied species is a combination of different species with only 1 recorded observation, this includes the clarki, scutalatus, tereticaudus, tigris, uniparens, and the viridis. 

3. What is the proportion of taxa included in this study? Show a table and plot that reflects this count.
Table: 
```{r}
deserts %>% 
  count(taxa) %>% 
  arrange(desc(n))
```
As shown above, rodents are the most studied taxa in this data. 

Plot: 
```{r}
deserts %>% 
  ggplot(aes(x=taxa,fill=taxa))+
  geom_bar()+
  scale_y_log10()+
  labs(title = "Studied Taxa",
       x= "Taxonomic Group",
       y= "Individuals")+
  theme(plot.title=element_text(size=rel(1.5),hjust=0.5))
```
Among the studied taxa, we can see rodents are the most abundantly studied. 

4. For the taxa included in the study, use the fill option to show the proportion of individuals sampled by `plot_type.`
```{r}
deserts %>% 
  ggplot(aes(x=species,fill=taxa))+
  geom_bar()+
  coord_flip()+
  scale_y_log10()+
  labs(title = "Studied Taxa",
       x= "Taxonomic Group",
       y= "Individuals")+
  theme(plot.title=element_text(size=rel(1.5),hjust=0.5))
```
While crowded, we can see the abundance of purple bars that represent the rodent taxa in comparison to the other taxa, indicating it to be the most highly studied taxa within the studied species. 

5. What is the range of weight for each species included in the study? Remove any observations of weight that are NA so they do not show up in the plot.

```{r}
deserts %>% 
  filter(!weight=="NA") %>% 
  ggplot(aes(x=weight, fill=species))+
  geom_boxplot(na.rm = T, position="dodge")+
  scale_y_log10()+
  labs(title = "Weight by Species",
       x= "Weight",
       y= "Species")+
  theme(plot.title=element_text(size=rel(1.5),hjust=0.5))
```

6. Add another layer to your answer from #5 using `geom_point` to get an idea of how many measurements were taken for each species.
```{r}
deserts %>% 
  filter(!weight=="NA") %>% 
  ggplot(aes(x=weight,y=species))+
  geom_point(na.rm = T)+
  labs(title = "Weight by Species",
       x= "Weight",
       y= "Species")+
  theme(plot.title=element_text(size=rel(1.5),hjust=0.5))
```

7. [Dipodomys merriami](https://en.wikipedia.org/wiki/Merriam's_kangaroo_rat) is the most frequently sampled animal in the study. How have the number of observations of this species changed over the years included in the study?

```{r}
names(deserts)
```


```{r}
deserts %>% 
  filter(species_id=="DM") %>% 
  group_by(year) %>% 
  summarise(n_samples=n()) %>% 
  ggplot(aes(x=as.factor(year),y=n_samples))+
    geom_col()+
  theme(axis.text.x = element_text(angle=45, hjust=1))+
  labs(title = "DM Observations per Year",
       x="Number of Observations",
       y= "Year")
```

8. What is the relationship between `weight` and `hindfoot` length? Consider whether or not over plotting is an issue.
```{r}
deserts %>% 
  ggplot(aes(x=weight,y=hindfoot_length))+
  geom_point(na.rm=T,size=1.2)+
  geom_smooth(method=lm,se=F)+
  scale_y_log10()+
   theme(axis.text.x = element_text(angle=45, hjust=1))+
  labs(title = "Weight and Hindfoot Length",
       x="Weight",
       y= "Hindfoot Length")
```
Over plotting is an issue with the data. Even with the the log10 scaling, the use of `jitter` instead of `point` to introduce random data into the plot, and making each plot point smaller each of the data points are still close together. Still, we are able to see a general pattern and relationship between the data points form. From this, we can see that hindfoot length will increase with weight, but when a weight of 100 is reached, corresponding hindfoot length will not increase. 

9. Which two species have, on average, the highest weight? Once you have identified them, make a new column that is a ratio of `weight` to `hindfoot_length`. Make a plot that shows the range of this new ratio and fill by sex.
```{r}
names(deserts)
```
```{r}
deserts %>% 
  group_by(species) %>% 
  filter(!weight=="NA") %>% 
  summarise(mean_weight=mean(weight,na.Rm=T)) %>% 
  arrange(desc(mean_weight))
```
Albigula and spectabilis are the species with the highest mean weight. 

Plot
```{r}
deserts %>% 
  filter(species_id=="DS" | species_id =="NL") %>% 
  filter(!weight=="NA" & !hindfoot_length=="NA" & !sex=="NA") %>% 
  mutate(ratio=weight/hindfoot_length) %>%
  select(ratio,species_id,weight,sex) %>% 
  ggplot(aes(x=ratio,fill=sex))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle=45, hjust=1))+
  labs(title = "Weight to Foot Ratios",
       x="Ratio")
```
Note: When specifying the NA exclusions from the data, we must use the `and` command and not the `or` command, because we only want to plot data that does not contain NA values for any of the 3 variables we are interested in plotting. Also, finding the range of variables corresponds to boxplots! 

10. Make one plot of your choice! Make sure to include at least two of the aesthetics options you have learned.
```{r}
names(deserts)
```
```{r}
options(scipen=999)
```

```{r}
deserts %>% 
 filter(taxa=="Rodent") %>% 
  filter(plot_type == "Control") %>% 
  filter(!sex== "NA") %>% 
  ggplot(aes(x=species,y=weight,fill=sex))+
  geom_boxplot(alpha=0.4)+
  scale_y_log10()+
  coord_flip()
```
There are multiple taxa in this data, but for some reason only `rodent` is the only taxa that is working with this code. Why? Shouldnt this code work with any taxa I want it to? 


## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 