---
title: "BIS15L_AI_hmwk"
author: "Jodie Cheun"
date: "`r Sys.Date()`"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
This homework is a bit less traditional this week. There are some errors in the code below. In addition add notes and titles to the code chunks that explains what the code chunck does. Use AI tools, notes from class, or any other resources to help fix the errors. Feel free to edit any plots to your preference. Don't forget to answer the two questions at the bottom. Upon completion, you should not have to '#' out any of the code chunks.

Push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.   

Again make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  


```{r}
library(tidyverse)
library(janitor)
library(dplyr)
```


1.
```{r}
blood_count <- read_csv("data/blood_count_dataset.csv") %>% clean_names()
```

2. 
```{r}
glimpse(blood_count)
```

3. 
```{r}
blood_count %>%
  group_by(gender) %>%
  summarize(min = min(red_blood_cells),
            mean = mean(red_blood_cells),
            max = max(red_blood_cells))
```


4. 
```{r}
options(scipen = 999, digits = 2)

# Assuming "Gender" is lowercase in your dataset
blood_count %>%
  filter(gender == "male") %>%  # Use double equals sign for equality check
  select(3:9) %>%  # Use parentheses instead of square brackets for select
  colMeans(na.rm = TRUE)  # Compute column means
```

5. 
```{r}
blood_count %>%
  filter(gender == "Female") %>%  # Use double equals sign for equality check
  select(3:9) %>%  # Use parentheses instead of square brackets for select
  colMeans(na.rm = TRUE)  # Compute column means
```


6. 
```{r}
blood_count %>%
  pivot_longer(cols = 3:9, names_to = "Test", values_to = "Values") %>%
  ggplot(aes(x = gender, y = Values, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~ Test, scales = "free") + 
  labs(title = "Boxplot of Blood Count Tests",
       x = "Test",
       y = "Values")
```

7. 
```{r}
grouped_bc <- blood_count %>%
  mutate(age_group = case_when(
    age >= 20 & age <= 29 ~ "20s",
    age >= 30 & age <= 39 ~ "30s",
    age >= 40 & age <= 49 ~ "40s",
    age >= 50 & age <= 59 ~ "50s",
    age >= 60 & age <= 70 ~ "60s")) # there were only 4 people in 70s, so I combined it with 60s

head(grouped_bc)
```

8. 
```{r}
grouped_bc %>%
  pivot_longer(cols = 3:9, names_to = "test", values_to = "values") %>%
  ggplot(aes(x = age_group, y = values, fill = age_group)) +
  geom_boxplot() +
  facet_wrap(~ test, scales = "free") +  # Facet by the "test" variable
  labs(title = "Boxplot of Blood Count Tests by Age Group",
       x = "Age Group",
       y = "Values",
       fill = "Age Group")  # Define legend title

```

9. What are two drawbacks and two benefits of AI?
Benefits:

Increased productivity: AI-powered tools can automate repetitive tasks such as code generation, debugging, and refactoring. This can significantly increase the productivity of developers by reducing the time spent on routine coding tasks, allowing them to focus on more complex and creative aspects of software development.

Improved code quality: AI tools can analyze code patterns, detect potential bugs, and suggest optimizations. By leveraging AI for code review and analysis, developers can identify and fix issues early in the development process, leading to higher-quality software with fewer bugs and vulnerabilities.

Drawbacks:

Overreliance on automation: While AI can automate many coding tasks, overreliance on AI-powered tools may lead to a decrease in developers' coding skills and critical thinking abilities. Relying too heavily on AI for coding tasks may hinder developers' ability to understand the underlying principles of programming and problem-solving.

Bias and errors: AI models used in coding tools are trained on historical data, which may contain biases and errors. This can result in AI tools making incorrect recommendations or perpetuating existing biases in the code. Developers need to be aware of the limitations of AI models and carefully evaluate the suggestions provided by AI-powered coding tools to avoid introducing unintended errors or biases into their code.

10. Do you think you will use AI for coding in the future? Why or why not?
These are the reasons provided by ChatGPT: 

Technological advancements: AI technology is continually improving, enabling more sophisticated applications in coding and software development.

Demand for automation: There's a growing need for automation to streamline workflows and increase efficiency, making AI-powered tools valuable for developers.

Data-driven decision-making: AI helps make sense of large volumes of data, providing insights that inform decision-making and improve software performance.

Enhanced developer productivity: AI-powered tools augment developers' capabilities, leading to faster development cycles and higher-quality software.

For the most part, I agree with these answers. The future is bound to become more automated and fast paced to keep up with human demand. AI will only assist humans (hopefully) in reaching goals. 

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 