---
title: "Homework 7"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
library(naniar)
``` 

## Data
**1. For this homework, we will use two different data sets. Please load `amniota` and `amphibio`.**  

`amniota` data:  
Myhrvold N, Baldridge E, Chan B, Sivam D, Freeman DL, Ernest SKM (2015). “An amniote life-history
database to perform comparative analyses with birds, mammals, and reptiles.” _Ecology_, *96*, 3109.
doi: 10.1890/15-0846.1 (URL: https://doi.org/10.1890/15-0846.1).
```{r}
amniota
```

```{r}
amniota <- clean_names(amniota)
```

`amphibio` data:  
Oliveira BF, São-Pedro VA, Santos-Barrera G, Penone C, Costa GC (2017). “AmphiBIO, a global database
for amphibian ecological traits.” _Scientific Data_, *4*, 170123. doi: 10.1038/sdata.2017.123 (URL:
https://doi.org/10.1038/sdata.2017.123).

```{r}
amphibio
```

```{r}
#amphibio <- clean_names(amphibio)
```


## Questions  
**2. Do some exploratory analysis of the `amniota` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  
```{r}
summary(amniota)
```

```{r}
names(amniota)
```
```{r}
amniota %>% 
  map_df(~sum(is.na(.)))
```
While the `map_df` does not show any NAs in the data, we can notice some strange observation values from the `summary` function, specifically the -999.0 values. As such, we need to fix the way NA are represented in the `amniota` data. 

**3. Do some exploratory analysis of the `amphibio` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  
```{r}
names(amphibio)
```
```{r}
summary(amphibio)
```
```{r}
amphibio %>% 
  map_df(~sum(is.na(.)))
```
As shown above, we see NA values labeled as `NA` and are hence recognized by the `map_df` function. From the `summary` function, I do not see any odd value placeholders used instead of NA. 

**4. How many total NA's are in each data set? Do these values make sense? Are NA's represented by values?**   

```{r}
amniota %>% 
  map_df(~sum(is.na(.)))
```

```{r}
miss_var_summary(amniota)
```
In the amniota dataset, neither the `map_df` or `miss_var_summary` functions report findings of NA in the data. However, we know this is suspicious from the -999.0 values observed by the `summary` function. Hinting these -999.0 values are likely placeholders for NA values in the data. 


```{r}
amphibio %>% 
  map_df(~sum(is.na(.)))
```
```{r}
miss_var_summary(amphibio) %>% 
  mutate(total_NA=sum(n_miss))
```

```{r}
amphibio %>% 
  summarise(total_NA=sum(is.na(amphibio)))
```

In the amphibio dataset, we observe a total of 170566 NA values from various variables. 
 
**5. Make any necessary replacements in the data such that all NA's appear as "NA".**   
```{r}
names(amniota)
```

```{r}
new_amniota <- amniota%>% 
  replace_with_na(replace = list(subspecies=-999,
                                 female_maturity_d=-999.000,
                                 litter_or_clutch_size_n=-999.000,
                                 litters_or_clutches_per_y=-999.0,
                                 maximum_longevity_y=-999.00000,
                                 gestation_d=-999,
                                 weaning_d=-999,
                                 birth_or_hatching_weight_g=-999.00,
                                 weaning_weight_g=-999,
                                 egg_mass_g=-999.00,
                                 incubation_d=-999.00,
                                 fledging_age_d=-999.00,
                                 longevity_y=-999.000000,
                                 male_maturity_d=-999.00,
                                 inter_litter_or_interbirth_interval_y= -999,
                                 female_body_mass_g=-999.000,
                                 male_body_mass_g=-999.0000,
                                 no_sex_body_mass_g=-999.000,
                                 egg_width_mm=-999.00, 
                                 egg_length_mm=-999.0,
                                 fledging_mass_g=-999.00,
                                 adult_svl_cm=-999.00,
                                 male_svl_cm=-999,
                                 female_svl_cm=-999,
                                 birth_or_hatching_svl_cm=-999,
                                 female_svl_cm=-999, 
                                 female_body_mass_at_maturity_g=-999,
                                 no_sex_svl_cm=-999,
                                 no_sex_maturity_d=-999,
                                 adult_body_mass_g=--30258.71))
new_amniota
```

```{r}
new_amniota %>% 
  summarise(total_na_amniota=sum(is.na(new_amniota)))
```

Since we do see accurately accounted for NA values in the amphibio data, no changes are needed there. However, in the amniota datset, we have several variations of placeholder values in almost all variables that need to be accounted for and changed to NA. Using the `replace_with_NA` function, we can specify the exact variation of the NA placeholder used in each variable spot. To double check out work, we can view the new version of the amniota data to check that all -999 iterations are removed and replaced with NA. 

**6. Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amniota` data.**  
```{r}
miss_var_summary(new_amniota)
```

**7. Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amphibio` data.**
```{r}
miss_var_summary(amphibio)
```

**8. For the `amniota` data, calculate the number of NAs in the `egg_mass_g` column sorted by taxonomic class; i.e. how many NA's are present in the `egg_mass_g` column in birds, mammals, and reptiles? Does this results make sense biologically? How do these results affect your interpretation of NA's?**  
```{r}
new_amniota %>% 
  group_by(class) %>% 
  select(class,egg_mass_g) %>% 
  miss_var_summary(order=T)
```
For the mammalia class, yes these results make sense biologically, simply because we know that mammals (except the platypus) do not lay eggs in their reproductive cycle. The aves and and Reptilia classes make less biological sense on why their taxonomc classes have such high proportions of NA values. While the Aves class do not have the worse percentage of NA values (although 50% is pretty high), the Reptilia class haveing almost 92% of missing data is quite concerning. A quick Google search to help explain this high percentage of NA values demonstrates that reptiles tend to be difficult to study/survey. 

**9. The `amphibio` data have variables that classify species as fossorial (burrowing), terrestrial, aquatic, or arboreal.Calculate the number of NA's in each of these variables. Do you think that the authors intend us to think that there are NA's in these columns or could they represent something else? Explain.**
```{r}
amphibio %>% 
  select(fos,ter,aqu,arb) %>% 
  miss_var_summary()
```

From the data itself, it seems a majority of the NA values are placed depending on the species being recorded. Specifically, all variables of fos, arb, aqu, and ter, are included and are makred 1 if this lifestyle is applicable to the species and marked NA if this lifestyle is not applicable to the species. For clarification, `lifestyle` is being used to decribe the types of lands each species depends/lives on. 

**10. Now that we know how NA's are represented in the `amniota` data, how would you load the data such that the values which represent NA's are automatically converted?**
```{r}
#new_amniota2 <- read_csv("data/amniota.csv",na=c("-999", "-999.0","-999.00", "-999.000","-999.0000","-999.00000", "-999.00000"))
```

After throughly going through the amniota data, we see a common pattern of -999 iterations being used as NA placeholders in various variables. As such, we can deal with these -999 values in advance by specifying all the values we want R to consider and replace with NA before the data gets loaded into R Studio for us to use. 
## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.  