---
title: "Homework 9"
author: "Jodie Cheun"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
``` 

For this homework, we will take a departure from biological data and use data about California colleges. These data are a subset of the national college scorecard (https://collegescorecard.ed.gov/data/). Load the `ca_college_data.csv` as a new object called `colleges`.
```{r}
colleges <- read_csv("data/ca_college_data.csv") %>% clean_names()
```

The variables are a bit hard to decipher, here is a key:  

INSTNM: Institution name  
CITY: California city  
STABBR: Location state  
ZIP: Zip code  
ADM_RATE: Admission rate  
SAT_AVG: SAT average score  
PCIP26: Percentage of degrees awarded in Biological And Biomedical Sciences  
COSTT4_A: Annual cost of attendance  
C150_4_POOLED: 4-year completion rate  
PFTFTUG1_EF: Percentage of undergraduate students who are first-time, full-time degree/certificate-seeking undergraduate students  

1. Use your preferred function(s) to have a look at the data and get an idea of its structure. Make sure you summarize NA's and determine whether or not the data are tidy. You may also consider dealing with any naming issues.

```{r}
names(colleges)
```

```{r}
str(colleges)
```
```{r}
colleges
```
Note that the colleges included in this data set are all in CA.  

```{r}
sum(is.na(colleges))
```

We have 949 NAs in the `colleges` data. 

2. Which cities in California have the highest number of colleges?
```{r}
colleges %>% 
  select(city,stabbr,instnm) %>% 
  filter(stabbr=="CA") %>% 
  group_by(city) %>% 
  summarise(n_colleges=n_distinct(instnm)) %>% 
  arrange(desc(n_colleges))
```
LA has the highest number of colleges with 24. 

3. Based on your answer to #2, make a plot that shows the number of colleges in the top 10 cities.
```{r}
colleges %>% 
  count(city) %>% 
  top_n(10,n) %>% 
  ggplot(aes(y=city, x=n))+
  geom_col()
```
Putting the cities on the y-axis (instead of the x-axis) makes it easier to read the different city names. 

4. The column `COSTT4_A` is the annual cost of each institution. Which city has the highest average cost? Where is it located?
```{r}
colleges %>%
  group_by(city,instnm) %>% 
  summarise(mean_cost=mean(costt4_a,na.rm=T)) %>% 
  arrange(desc(mean_cost))
```
The Harvey Mudd College located in Claremont city had the highest mean annual cost of attendance at 66498. This city is located in CA. 

5. Based on your answer to #4, make a plot that compares the cost of the individual colleges in the most expensive city. Bonus! Add UC Davis here to see how it compares :>).

Code alone: 
```{r}
colleges %>% 
  filter(city=="Claremont") %>% 
  group_by(city, instnm) %>% 
  summarise(total_cost=sum(costt4_a,rm.na=T)) %>% 
  arrange(desc(total_cost)) 
```
Resulting plot (and adding UCD): 
```{r}
colleges %>% 
  filter(city=="Claremont") %>% 
  group_by(city, instnm) %>% 
  summarise(total_cost=sum(costt4_a,rm.na=T)) %>% 
  arrange(desc(total_cost)) %>% 
  ggplot(aes(y=instnm,x=total_cost))+
  geom_col(,na.rm = T)
```
Note to self: Missing the inclusion of Davis in the above data^^

6. The column `ADM_RATE` is the admissions rate by college and `C150_4_POOLED` is the four-year completion rate. Use a scatterplot to show the relationship between these two variables. What do you think this means?
```{r}
colleges %>% 
  ggplot(aes(x=adm_rate,y=c150_4_pooled))+
  geom_point()+
  geom_smooth(method = lm,se=T)
```
In the plot shown above, we see there is a negative relationship between college admission rate and 4 year completion rate. The higher the admission rate, the lower the overall 4 year completion rate. On the other hand, the lower the admission rate, the higher the 4 year completion rate. 


7. Is there a relationship between cost and four-year completion rate? (You don't need to do the stats, just produce a plot). What do you think this means?
```{r}
colleges %>% 
  ggplot(aes(x=adm_rate,y=costt4_a))+
  geom_point()+
  geom_smooth(method = lm,se=T)
```
No, there does not seem to be a relationship between admission rate and cost of attendance. As shown in the plot above, the data points between these axis are highly variable with no clear increasing or decreasing trend. 

8. The column titled `INSTNM` is the institution name. We are only interested in the University of California colleges. Make a new data frame that is restricted to UC institutions. You can remove `Hastings College of Law` and `UC San Francisco` as we are only interested in undergraduate institutions.
```{r}
uc_colleges <- colleges %>% 
  select(instnm,city,stabbr,zip,adm_rate,sat_avg,pcip26,costt4_a,c150_4_pooled,pftftug1_ef) %>% 
  filter(instnm=="University of California-San Diego" | instnm=="University of California-Irvine" | instnm== "University of California-Riverside" | instnm== "University of California-Los Angeles" | instnm =="University of California-Davis" | instnm=="University of California-Santa Cruz" |instnm== "University of California-Santa Barbara" | instnm== "University of California-Berkeley")
uc_colleges
```

Remove `Hastings College of Law` and `UC San Francisco` and store the final data frame as a new object `univ_calif_final`.
```{r}
univ_calif_final<- uc_colleges 
```
Since I manually selected which colleges to include in the filter function, the `Hastings College of Law` and `UC San Francisco` colleges are not included, and hence do not to be removed from the pre-existing `uc_colleges` object. 

Use `separate()` to separate institution name into two new columns "UNIV" and "CAMPUS".
```{r}
univ_calif_final %>% 
  separate(instnm,into = c("univ","campus"),sep="-")
```

9. The column `ADM_RATE` is the admissions rate by campus. Which UC has the lowest and highest admissions rates? Produce a numerical summary and an appropriate plot.
```{r}
univ_calif_final %>% 
  separate(instnm,into = c("univ","campus"),sep="-") %>% 
  group_by(campus) %>% 
  summarise(total_admission=sum(adm_rate)) %>% 
  arrange(desc(total_admission))
```
From the numerical summary shown above, the campus with the highest admission rate is UC Riverside, and the campus with the lowest admission rate is UC Berkeley. 

```{r}
univ_calif_final %>% 
  separate(instnm,into = c("univ","campus"),sep="-") %>% 
  group_by(campus) %>% 
  summarise(total_admission=sum(adm_rate)) %>% 
  arrange(desc(total_admission)) %>% 
  ggplot(aes(y=campus,x=total_admission)) +
  geom_col()
```

10. If you wanted to get a degree in biological or biomedical sciences, which campus confers the majority of these degrees? Produce a numerical summary and an appropriate plot. (Using the PCIP26 vairable)
```{r}
univ_calif_final %>% 
  separate(instnm,into = c("univ","campus"),sep="-") %>% 
  group_by(campus) %>% 
  summarise(total_bio_degrees=sum(pcip26)) %>% 
  arrange(desc(total_bio_degrees))
```
From the numerical summary above, we see that UC San Diego, of the UCs, has the highest majority of biological or biomedical science degrees. 

```{r}
univ_calif_final %>% 
  separate(instnm,into = c("univ","campus"),sep="-") %>% 
  group_by(campus) %>% 
  summarise(total_bio_degrees=sum(pcip26)) %>% 
  ggplot(aes(y=campus,x=total_bio_degrees))+
  geom_col()
```


## Knit Your Output and Post to [GitHub](https://github.com/FRS417-DataScienceBiologists)