---
title: "BIS15L_AI_class"
author: "Jodie Cheun"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
Today, we're exploring how AI can aid in writing, comprehending and troubleshooting our R code. AI is a very valuable tool, but it's important not to rely on it entirely. Building a strong R foundation is still important. These AI tools are not perfect and can be inaccurate solutions. In addition these tools can't read our minds, so understanding R helps us generate better prompts to get better responses. In short, AI should be used as a tool, not a crutch. Additionally, do not forget about other useful resources like Stack Overflow/Exchange, online tutorials, and GitHub issues pages for learning and troubleshooting. Let's combine AI with traditional methods to master R effectively.

This overview isn't exhaustive of all possible AI tools for R coding, but I believe it is a good start and covers everything you need.


**Make sure to enable Github Copilot in RStudio.**
Go to Tools -> Global Options -> Copilot -> Enable Git

## Learning Goals: 
*At the end of this exercise, you will be able to:*  
1. Explain what AI is and what AI tools can be used for R.
2. Enable Github Copilot in RStudio.
3. Use Copilot/ChatGPT to write code in RStudio.
4. Use Copilot/ChatGPT for troubleshooting code in RStudio.


## Exploration Questions:
Lets get some get a feel for what copilot (or ChatGPT) does, use Github Copilot to autocomplete the questions below. You can alternatively paste the questions in ChatGPT or Microsoft Copilot.
1. What is AI?

2. What is ChatGPT?

3. What is Github Copilot?

4. [Try it yourself: Insert your own question here]. What is the most useful library to use in R. 

## Load the libraries  
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(dplyr)
library(shiny)
library(shinydashboard)
```

## Load the Data: 
```{r}
#load the Netflix Data in your Data folder, save it under the object name netflix

netflix <- read_csv("Data/netflix_titles.csv")

```

*Did copilot help you write the code above? Could anything about the prompt be improved to help Copilot understand what you want to do more accurately?*


## Lets Practice:
Explore the elements in your data: include a comment within your code chunk to prompt Copilot to use the function of your choice
```{r}
head(netflix)

```
*Did copilot use your preferred method to get the summary of the data? If not, how could you improve the prompt to get the desired output?*

```{r}
names(netflix)
```


Sort out the movies from the netflix into an new object called netflix_movies
```{r} 
netflix_movies <- filter(netflix,type=="Movie")
```


Which director has the most movies on netflix?
```{r}
library(dplyr)

# Assuming netflix_movies is your dataframe containing Netflix movies data

netflix_movies %>%
  count(director, sort = TRUE) %>%
  filter(!is.na(director))

```
*Do you see anything wrong with the table above, look at row 1. Write a prompt below to fix this.*

[insert your prompt here]
```{r}
library(dplyr)
library(stringr)

# Assuming netflix_movies is your dataframe containing Netflix movies data

netflix_movies <- netflix_movies %>%
  filter(str_count(director, ",") < 1)
netflix_movies
```

```{r}
netflix_movies %>%
  count(director, sort = TRUE) %>%
  filter(!is.na(director))
```


Which director had the most movies by release year and remove the NA's?
```{r}
library(dplyr)

# Assuming netflix_movies is your dataframe containing Netflix movies data

# Remove rows with missing values in the director column
netflix_movies <- netflix_movies %>%
  filter(!is.na(director))

# Group by release year and director, count movies, and select the top director per year
top_director_per_year <- netflix_movies %>%
  group_by(release_year, director) %>%
  count() %>%
  top_n(1, wt = n) %>%
  ungroup()

# View the top director per year
print(top_director_per_year)


```
*See anything else interesting about this table, look at the first director on the table. Could this be a problem and how can we edit the dataset to fix this? Try writing a prompt.*

Which director has the most movies on netflix? Remove the NAs
```{r}
netflix_movies %>% 
  count(netflix_movies, director, sort=T,!is.na(director))
```

Make a bar plot of the top directors with the most movies on netflix
```{r}
library(dplyr)
library(ggplot2)

# Assuming your dataframe is named 'netflix_movies'

# Count the number of movies per director
director_counts <- netflix_movies %>%
  group_by(director) %>%
  summarise(num_movies = n()) %>%
  arrange(desc(num_movies)) %>%
  top_n(10)  # Select top 10 directors with the most movies

# Create a bar plot with sage green bars and black outline
ggplot(director_counts, aes(x = reorder(director, num_movies), y = num_movies)) +
  geom_bar(stat = "identity", fill = "#9DC183", color = "black") +  # Sage green color using hexadecimal code
  labs(title = "Top Directors with Most Movies on Netflix",
       x = "Director",
       y = "Number of Movies") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
*Do you see any issues with this plot? How could we improve the prompt to get a better boxplot? Try writing a prompt.*

[Insert your prompt here]
```{r}

```

*How about for just 2020*

[Insert your prompt here]
```{r}
library(dplyr)
library(ggplot2)

# Assuming your dataframe is named 'netflix_movies'

# Filter data for the year 2020
netflix_movies_2020 <- netflix_movies %>%
  filter(substr(release_year, 1, 4) == "2020")

# Count the number of movies per director in 2020
director_counts_2020 <- netflix_movies_2020 %>%
  group_by(director) %>%
  summarise(num_movies = n()) %>%
  arrange(desc(num_movies)) %>%
  top_n(10)  # Select top 10 directors with the most movies in 2020

# Create a bar plot for directors with most movies in 2020
ggplot(director_counts_2020, aes(x = reorder(director, num_movies), y = num_movies)) +
  geom_bar(stat = "identity", fill = "#9DC183", color = "black") +  # Sage green color with black outline
  labs(title = "Top Directors with Most Movies on Netflix in 2020",
       x = "Director",
       y = "Number of Movies") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))  # Center the plot title

```


*Take a few minutes to make your own plot(s) using prompts, edit prompts until you get a plot that you like.*

## Lets make the top 10 directors barplot into a Shiny App: use copilot or Chatgpt
```{r}
library(shiny)
library(shinydashboard)
library(dplyr)
library(ggplot2)

# Define UI for the Shiny app
ui <- dashboardPage(
  dashboardHeader(title = div("Movie Explorer", style = "font-family: 'Times New Roman';")),
  dashboardSidebar(
    sidebarMenu(
      menuItem("Explore Movies", tabName = "explore", icon = icon("film"))
    )
  ),
  dashboardBody(
    tabItems(
      tabItem(
        tabName = "explore",
        fluidRow(
          box(
            title = "Number of Movies per Director",
            plotOutput("director_plot")
          )
        ),
        fluidRow(
          box(
            title = "Select Year",
            selectInput("year", "Select Year:",
                        choices = unique(substr(netflix_movies$release_year, 1, 4)),
                        selected = "2020")
          )
        ),
        fluidRow(
          box(
            title = "Movie Details",
            tableOutput("movie_table")
          )
        )
      )
    )
  ),
  skin = "blue"  # Change the skin color of the dashboard
)

# Define server logic for the Shiny app
server <- function(input, output) {
  
  # Filter data based on selected year
  selected_movies <- reactive({
    selected_year <- input$year
    filter(netflix_movies, substr(release_year, 1, 4) == selected_year)
  })
  
  output$movie_table <- renderTable({
    # Display movie details with only type, title, and director
    selected_movies() %>%
      select(type, title, director)
  })
  
  output$director_plot <- renderPlot({
    # Plot number of movies per director with sage green bars and centered title
    ggplot(selected_movies(), aes(x = director)) +
      geom_bar(fill = "#9DC183") +  # Sage green color
      labs(title = "Number of Movies per Director",
           x = "Director",
           y = "Number of Movies") +
      theme_minimal() +
      theme(plot.title = element_text(family = "Times New Roman"))  # Change font to Times New Roman
  })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)

```



*Now use ChatGPT, Copilot or previous notes to fix the code*





## That's it! Let's take a break and then move on to working on group projects.
