---
title: "Midterm 2 W24"
author: "Jodie Cheun"
date: "`r Sys.Date()`"
output:
  html_document: 
    keep_md: yes
  pdf_document: default
---

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code must be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Be sure to add your name to the author header above. 

Your code must knit in order to be considered. If you are stuck and cannot answer a question, then comment out your code and knit the document. You may use your notes, labs, and homework to help you complete this exam. Do not use any other resources- including AI assistance.  

Don't forget to answer any questions that are asked in the prompt. Some questions will require a plot, but others do not- make sure to read each question carefully.  

For the questions that require a plot, make sure to have clearly labeled axes and a title. Keep your plots clean and professional-looking, but you are free to add color and other aesthetics.  

Be sure to follow the directions and upload your exam on Gradescope.    

## Background
In the `data` folder, you will find data about shark incidents in California between 1950-2022. The [data](https://catalog.data.gov/dataset/shark-incident-database-california-56167) are from: State of California- Shark Incident Database.   

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("naniar")
```

## Load the data
Run the following code chunk to import the data.
```{r message=FALSE, warning=FALSE}
sharks <- read_csv("data/SharkIncidents_1950_2022_220302.csv") %>% clean_names()
```

## Questions
1. (1 point) Start by doing some data exploration using your preferred function(s). What is the structure of the data? Where are the missing values and how are they represented?  
```{r}
names(sharks)
```

```{r}
glimpse(sharks)
```
```{r}
miss_var_summary(sharks)
```

```{r}
summary(sharks)
```
Given we do not see any extreme values in the min and max, there are no number placeholders for NA. We do see placeholders in the forms of `not counted` and `unknown` that will have to be accounted for. 

2. (1 point) Notice that there are some incidents identified as "NOT COUNTED". These should be removed from the data because they were either not sharks, unverified, or were provoked. It's OK to replace the `sharks` object.

```{r}
names(sharks)
```

```{r}
sharks <- sharks %>% 
  filter(!incident_num == "NOT COUNTED") %>% 
  filter(!time == "Unknown") %>% 
  filter(!species == "Unknown")
#sharks
```

3. (3 points) Are there any "hotspots" for shark incidents in California? Make a plot that shows the total number of incidents per county. Which county has the highest number of incidents?

*Data*
```{r}
sharks %>% 
  select(location,county,injury,incident_num) %>% 
  group_by(county) %>% 
  summarise(n_incidents=n_distinct(incident_num)) %>% 
  arrange(desc(n_incidents))
```
We can see that the Humboldt, Santa Barbara, and San Diego county are "hotspots" for shark incidents, and has 17 distinct incidents marked in the data. 
 
*Plot*
```{r}
sharks %>% 
  select(location,county,injury,incident_num) %>% 
  group_by(location,county) %>% 
  summarise(n_incidents=n_distinct(incident_num)) %>% 
  arrange(desc(n_incidents)) %>% 
  ggplot(aes(x=n_incidents,y=county,fill=county))+
  geom_col(na.rm=T,alpha=0.6)+
  labs(title="Shark Incidents Per County",
      x= "Number of Incidents",
      y= "County")+
  theme_light()
```
4. (3 points) Are there months of the year when incidents are more likely to occur? Make a plot that shows the total number of incidents by month. Which month has the highest number of incidents?  

*Data*
```{r}
sharks %>% 
  group_by(month) %>% 
  summarise(n_incidents=n_distinct(incident_num))
```
Month 10 (aka October) is recorded to have the most distinct counts of incidents at 32. Yikes. 
*Need to convert the month to a factor so R stops treating them as numbers* 

*Plot*
```{r}
sharks %>% 
  group_by(month) %>% 
  summarise(n_incidents=n_distinct(incident_num)) %>% 
  ggplot(aes(x=month,y=n_incidents,fill=month))+
  geom_col(color="black",alpha=0.6)+
  labs(title= "Shark Incidents by Month",
       x= "Months",
       y="Number of Shark Incidents")+
  theme_get()
```

5. (3 points) How do the number and types of injuries compare by county? Make a table (not a plot) that shows the number of injury types by county. Which county has the highest number of fatalities?

```{r}
sharks %>% 
  select(county,incident_num,injury) %>% 
  filter(injury =="fatal") %>% 
  group_by(county) %>% 
  count(injury) %>% 
  arrange(desc(n))
```
The San Luis Obispo county has the most number of injuries categorized as "fatal". 

6. (2 points) In the data, `mode` refers to a type of activity. Which activity is associated with the highest number of incidents?

```{r}
sharks %>% 
  select(county,injury,mode,incident_num) %>% 
  filter(injury!="none") %>% 
  group_by(mode) %>% 
  count(injury) %>% 
  arrange(desc(n))
```
Surfing and boarding had the highest number of incidents with injuries labeled as either minor or major. 

7. (4 points) Use faceting to make a plot that compares the number and types of injuries by activity. (hint: the x axes should be the type of injury) 
```{r}
sharks %>% 
  select(county,injury,mode,incident_num) %>% 
  filter(injury!="none") %>% 
  group_by(mode) %>% 
  count(injury) %>% 
  ggplot(aes(x=injury,y=n,fill=injury))+
  geom_col(color="black",alpha=0.6)+
  facet_wrap(~mode)+
  theme(axis.text.x=element_text(angle=60,hjust = 1))+
  labs(title="Injuries per Activity",
       x=" Injury Type",
       y="Number of Injuries")
```

8. (1 point) Which shark species is involved in the highest number of incidents?  
```{r}
sharks %>% 
  select(species,incident_num) %>% 
  filter(species!="NA") %>% 
  filter(species!="Unknown") %>% 
  filter(incident_num!="NA") %>% 
  group_by(species) %>% 
  summarise(total_incidents=n_distinct(incident_num)) %>% 
  arrange(desc(total_incidents))
```
By a landslide, the White shark species has the highest total number of incidents. 

9. (3 points) Are all incidents involving Great White's fatal? Make a plot that shows the number and types of injuries for Great White's only.  
```{r}
sharks %>% 
  select(species, incident_num, injury) %>% 
  filter(species=="White") %>% 
  filter(injury!="NA") %>% 
  group_by(injury) %>%
  summarise(incident_total=n_distinct(incident_num)) %>% 
  arrange(desc(incident_total)) %>% 
  ggplot(aes(x=injury,y=incident_total,fill=injury))+
  geom_col(color="black",alpha=0.6)+
  labs(title = "Great White Injuries and Incidents",
       x="Injury Type",
       y="Number of Incidents by Great Whites")+
  theme_grey()
```
No, not all the injuries from the Great White are considered fatal. 

## Background
Let's learn a little bit more about Great White sharks by looking at a small dataset that tracked 20 Great White's in the Fallaron Islands. The [data](https://link.springer.com/article/10.1007/s00227-007-0739-4) are from: Weng et al. (2007) Migration and habitat of white sharks (_Carcharodon carcharias_) in the eastern Pacific Ocean.

## Load the data
```{r message=FALSE, warning=FALSE}
white_sharks <- read_csv("data/White sharks tracked from Southeast Farallon Island, CA, USA, 1999 2004.csv", na = c("?", "n/a")) %>% clean_names()
```

10. (1 point) Start by doing some data exploration using your preferred function(s). What is the structure of the data? Where are the missing values and how are they represented?
```{r}
names(white_sharks)
```

```{r}
glimpse(white_sharks)
```
```{r}
miss_var_summary(white_sharks)
```
It seems that all NAs are accounted for and not hidden with placeholders. 

11. (3 points) How do male and female sharks compare in terms of total length? Are males or females larger on average? Do a quick search online to verify your findings. (hint: this is a table, not a plot).  

```{r}
white_sharks %>% 
  select(shark,sex,total_length_cm) %>% 
  filter(sex!="NA") %>% 
  group_by(sex) %>% 
  summarise(mean_length=mean(total_length_cm))
```
As shown above, the female sharks, on average, have a larger length. This is confirmed by Google! 

12. (3 points) Make a plot that compares the range of total length by sex. 
```{r}
white_sharks %>% 
  select(shark,sex,total_length_cm) %>% 
  filter(sex!="NA") %>% 
  group_by(sex) %>% 
  summarise(mean_length=mean(total_length_cm)) %>% 
  ggplot(aes(x=mean_length,y=sex,fill=sex))+
  geom_col(color="black",alpha=0.6)+
  labs(title="Mean Length by Sex",
       x="Total Mean Length (cm)",
       y="Sex")+
  theme_bw()
```

13. (2 points) Using the `sharks` or the `white_sharks` data, what is one question that you are interested in exploring? Write the question and answer it using a plot or table. 

*My question* From the `sharks` data, which county had the greatest number of White shark incidents? 

*Data*
```{r}
sharks %>% 
  select(county,location,incident_num,species) %>% 
  filter(species=="White") %>% 
  group_by(county) %>%
  summarise(total_incidents=n_distinct(incident_num)) %>% 
  arrange(desc(total_incidents))
```

*Plot*
```{r}
sharks %>% 
  select(county,location,incident_num,species) %>% 
  filter(species=="White") %>% 
  group_by(county) %>%
  summarise(total_incidents=n_distinct(incident_num)) %>% 
  arrange(desc(total_incidents)) %>% 
  ggplot(aes(x=county,y=total_incidents,fill=county))+
  geom_col(color="black",alpha=0.6)+
  labs(title="Great White Incidents by County",
       x="County",
       y="Number of White Incidents")+
  theme(axis.text.x=element_text(angle=45,hjust=1))
```
If we are just interested in the number of incidents caused by Great Whites in the data, Humboldt is the county that has the highest number of incidents. 
