---
title: "Tidyr 2: `pivot_wider()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Explain the difference between tidy and messy data.  
2. Evaluate a data set as tidy or untidy.  
3. Use the `pivot_wider()` function of `tidyr` to transform data from long to wide format.  
 
## Resources  (look into this!! )
- [tidyr `pivot_wider()`](https://tidyr.tidyverse.org/reference/pivot_wider.html)  
- [pivoting](https://cran.r-project.org/web/packages/tidyr/vignettes/pivot.html)  
- [R Ladies Sydney](https://rladiessydney.org/courses/ryouwithme/02-cleanitup-5/)  

## Review
Last time we learned the fundamentals of tidy data and used the `pivot_longer()` function to wrangle a few examples of frequently encountered untidy data. In the second part of today's lab we will look at a few more examples of `pivot_longer()` but also use the `pivot_wider()` function to deal with another type of untidy data.  

## Load the tidyverse
```{r}
library("tidyverse")
```

## `pivot_longer()`
Recall that we use `pivot_longer()` when our column names actually represent variables. A classic example would be that the column names represent observations of a variable.
```{r}
datasets::USPersonalExpenditure
?USPersonalExpenditure
```
Here we have data being represented as column names, bad. Will not be able to make a chart from this. 

Here we add a new column of expenditure types, which are stored as rownames above, with `mutate()`. The `USPersonalExpenditures` data also needs to be converted to a data frame before we can use the tidyverse functions, because it comes as a matrix.
```{r}
expenditures <- USPersonalExpenditure %>% 
  as_tibble() %>% #this transforms the matrix into a data frame
  mutate(expenditure = rownames(USPersonalExpenditure))
expenditures
```

## Practice
1. Are these data tidy? Please use `pivot_longer()` to tidy the data.
```{r}
expenditures %>% 
  pivot_longer(-expenditure, #aka do not move the expenditure 
               names_to = "year", #making the new variable year 
               values_to = "bn_dollars") #moving the values into the bn_dollars 
```
In this new form, now we have a year (x) and bn_dollars (y) 

2. Restrict the data to private education only. Sort in ascending order.
```{r}
expenditures %>% 
  pivot_longer(-expenditure,
               names_to = "year",
               values_to="bn_dollars") %>% 
  filter(expenditure=="private education")
  
```

## `separate()`
In this new heart rate example, we have the sex of each patient included with their name. Are these data tidy? No, there is more than one value per cell in the patient column and the columns a, b, c, d once again represent values.
```{r}
heartrate2 <- read_csv("data/heartrate2.csv")
heartrate2
```
Have 2 observations in the same cell and observations as columns 

We need to start by separating the patient names from their sexes. `separate()` needs to know which column you want to split, the names of the new columns, and what to look for in terms of breaks in the data.
```{r}
heartrate2 %>% 
  separate(patient, into= c("patient", "sex"), sep = "_")
```
Separating the pre-existing patient variable into separate variables of patient and sex. Where the values ae seperated by the presence of the _ 

## Practice
1. Re-examine `heartrate2`. Use `separate()` for the sexes, `pivot_longer()` to tidy, and `arrange()` to organize by patient and drug. Store this as a new object `heartrate3`.  
```{r}
heartrate3 <- heartrate2 %>% 
  separate(patient, into=c("patient", "sex"), sep="_") %>% 
  pivot_longer(-c(patient, sex),
               names_to = "drug",
               values_to = "heartrate")
heartrate3
```

2. `unite()` is the opposite of separate(). Its syntax is straightforward. You only need to give a new column name and then list the columns to combine with a separation character.  Give it a try below by recombining patient and sex from `heartrate3`.  
```{r}
heartrate3 %>% 
  unite(patient_sex, "patient", "sex", sep=" ")
```
Unite takes the seperation of data and puts observations back into the samevariable 

## `pivot_wider()`
The opposite of `pivot_longer()`. You use `pivot_wider()` when you have an observation scattered across multiple rows. In the example below, `cases` and `population` represent variable names not observations.  

Rules:  
+ `pivot_wider`(names_from, values_from)  
+ `names_from` - Values in the `names_from` column will become new column names  
+ `values_from` - Cell values will be taken from the `values_from` column  

```{r}
tb_data <- read_csv("data/tb_data.csv")
tb_data
```
Here, the value column is not represented of a single variable 
When using `pivot_wider()` we use `names_from` to identify the variables (new column names) and `values_from` to identify the values associated with the new columns.
```{r}
tb_data %>% 
  pivot_wider(names_from = "key", #taking the observations under key will become new columns
              values_from = "value") #the values under value will be moved to the new columns
```
pivot_wider is the opposite of pivot_longer. Here we are going from long to wide 

## Practice
1. Load the `gene_exp.csv` data as a new object `gene_exp`. Are these data tidy? Use `pivot_wider()` to tidy the data.
```{r}
gene_exp <- read_csv("data/gene_exp.csv")
gene_exp
```
More than one variable represented in each column. We can not hae treatment and control in te same observation

```{r}
gene_exp %>% 
  pivot_wider(names_from = "type",
              values_from = "L4_values")
```

## Practice
For the last practice example, I will use data from the awesome [R Ladies Sydney](https://rladiessydney.org/courses/ryouwithme/02-cleanitup-5/) blog. This data set is compiled by the NSW Office of Environment and Heritage contains the enterococci counts in water samples obtained from Sydney beaches as part of the Beachwatch Water Quality Program! The data set we’ll be working with is current as of October 13th 2018.  

1. Load the beachbugs data and have a look.
```{r}
beachbugs <- read_csv("data/beachbugs_long.csv")
beachbugs
```

2. Use `pivot_wider` to transform the data into wide format.
```{r}
beachbugs_wide <- beachbugs %>% 
  pivot_wider(names_from = site,
              values_from = buglevels)
beachbugs_wide
```
This is an overall better way to present the data, to show how beach contamination changes per year 
3. Now, use `pivot_longer` to transform them back to long!
```{r}
beachbugs_wide %>% 
  pivot_longer(-year,
               names_to = "site",
               values_to = "buglevels")
```

## Wrap-up  
Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)