---
title: "Lab_6_Warmup"
output: 
  html_document: 
    keep_md: yes
date: "2024-01-30"
---

##Load the libraries 
```{r}
library(tidyverse)
library(janitor)
```

##1. Load the bison.csv data. 
```{r}
bison <- read.csv("data/bison.csv")
```

##2. What are the dimensions and structure of the data? 
```{r}
glimpse(bison)
```
dbl stands for a numeric value 

```{r}
dim(bison)
```

```{r}
str(bison)
```
##3. We are only interested in code, sex, weight, year of birth. Restrict the data to these variables and store the dataframe as a new object. 
```{r}
summary(bison)
```
```{r}
bison %>% 
  select(animal_code, animal_sex, animal_weight, animal_yob)## Alt version of code with piping 
```


```{r}
new_bison <- select(bison, "animal_code","animal_sex", "animal_weight", "animal_yob")
new_bison
```

##Pull out the animals born between 1980- 1990 
```{r}
new_bison %>% 
  filter(animal_yob >= 1980 & animal_yob <=1990)
```
##How many male and female bison are represented between 1980 to 1990
```{r}
male_bison <- new_bison %>% 
  filter(animal_yob >= 1980 & animal_yob <=1990) %>% 
  filter(animal_sex=="M")
```

```{r}
table(male_bison$animal_sex)
```

```{r}
female_bison <- new_bison %>% 
  filter(animal_yob >= 1980 & animal_yob <=1990) %>% 
  filter(animal_sex=="F")
```

```{r}
table(female_bison$animal_sex)
```

##Mean of males 
```{r}
mean(male_bison$animal_weight)
```

##Mean of females 
```{r}
mean(female_bison$animal_weight)
```

