---
title: "dplyr Superhero"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    toc: true
    toc_float: true
    keep_md: true
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. Develop your dplyr superpowers so you can easily and confidently manipulate dataframes.  
2. Learn helpful new functions that are part of the `janitor` package.  

## Instructions
For the second part of lab today, we are going to spend time practicing the dplyr functions we have learned and add a few new ones. This lab doubles as your homework. Please complete the lab and push your final code to GitHub.  

## Load the libraries
```{r}
library("tidyverse")
library("janitor")
```

## Load the superhero data
These are data taken from comic books and assembled by fans. The include a good mix of categorical and continuous data.  Data taken from: https://www.kaggle.com/claudiodavi/superhero-set  

Check out the way I am loading these data. If I know there are NAs, I can take care of them at the beginning. But, we should do this very cautiously. At times it is better to keep the original columns and data intact.  
```{r}
superhero_info <- read_csv("data/heroes_information.csv", na = c("", "-99", "-"))
superhero_powers <- read_csv("data/super_hero_powers.csv", na = c("", "-99", "-"))
```
In the read call, we specify the values we want R to interpret as NA. Do not add this function to your read calls until you really understand the data 
## Data tidy
1. Some of the names used in the `superhero_info` data are problematic so you should rename them here. Before you do anything, first have a look at the names of the variables. You can use `rename()` or `clean_names()`.    
```{r}
superhero_info <- clean_names(superhero_info)
superhero_info
```

## `tabyl`
The `janitor` package has many awesome functions that we will explore. Here is its version of `table` which not only produces counts but also percentages. Very handy! Let's use it to explore the proportion of good guys and bad guys in the `superhero_info` data.  
```{r}
tabyl(superhero_info, alignment)
```

1. Who are the publishers of the superheros? Show the proportion of superheros from each publisher. Which publisher has the highest number of superheros?  

```{r}
summary(superhero_info)
```
```{r}
summary(superhero_powers)
```

```{r}
superhero_info %>% 
  tabyl(publisher) %>% 
  arrange(desc(percent))
```
With a combination of `tabyl` and `desc` to specify the data we want from the dataframe, we observe that Marvel Comics publisher has the highest percentage of superheros (n-value). 

2. Notice that we have some neutral superheros! Who are they? List their names below.  
```{r}
superhero_info %>% 
  select(name, alignment) %>% 
  filter(alignment=="neutral")
```
Notice that because we specify the dataframe used in the first line, we do no thave to include it with the `select` command in the piping steps afterwards. 

## `superhero_info`
3. Let's say we are only interested in the variables name, alignment, and "race". How would you isolate these variables from `superhero_info`?
```{r}
names(superhero_info)
```

```{r}
superhero_info %>% 
  select(name,alignment,race)
```

## Not Human
4. List all of the superheros that are not human.
```{r}
superhero_info %>% 
  filter(race != "Human")
```
Please note that only superheros with `Human` listed in the `race` column were considered `Human` for this analysis. As such, human hybride superheros were considered `Not Human`

## Good and Evil
5. Let's make two different data frames, one focused on the "good guys" and another focused on the "bad guys".
```{r}
good_guys <- (superhero_info) %>% 
  select(name, alignment, race) %>% 
  filter(alignment== "good")
good_guys
```

```{r}
bad_guys <- (superhero_info) %>% 
  select(name, alignment, race) %>% 
  filter(alignment== "bad")
bad_guys
```

6. For the good guys, use the `tabyl` function to summarize their "race".
```{r}
good_guys <- (superhero_info) %>% 
  select(name, alignment, race) %>% 
  filter(alignment== "good") %>% 
  tabyl(race) %>% 
  arrange(desc(percent))
good_guys
```

7. Among the good guys, Who are the Vampires?
```{r}
good_guys_vampire <- select(superhero_info, name, alignment,race) %>% 
  filter(alignment== "good") %>% 
  filter(race=="Vampire")

good_guys_vampire
```
As shown, Angel and Blade are the only Vampires with a `good` alignment.

8. Among the bad guys, who are the male humans over 200 inches in height?
```{r}
tall_bad_guys <- select(superhero_info, name, gender,alignment, height, race) %>% 
  filter(alignment == "bad") %>% 
  filter(gender== "Male") %>% 
  filter(race=="Human") %>% 
  filter(height> 200)

tall_bad_guys
```
As shown, we have 5 bad guys with a height over 200 (no units provided)

9. Are there more good guys or bad guys with green hair?  
```{r}
superhero_hair_colors <- select(superhero_info, name, alignment, hair_color) %>% 
  filter(hair_color=="Green")
superhero_hair_colors
```
By making a new dataframe specifically containing only the variables name, alignment, and hair color we can drastically reduce the observations to analyze. Further reduction is done by filtering observations in the `hair_color` column to only include superheros with `Green`. From this, we can conclude there are more good guys with green hair. 

10. Let's explore who the really small superheros are. In the `superhero_info` data, which have a weight less than 50? Be sure to sort your results by weight lowest to highest.  
```{r}
small_superheros <- select(superhero_info, name, alignment, weight) %>% 
  filter(weight <50) %>% 
  arrange(weight)
small_superheros
```
As shown, Iron Monger is the smallest superhero with a weight of 2 (no units specified), and they are a bad guy! 

11. Let's make a new variable that is the ratio of height to weight. Call this variable `height_weight_ratio`.    
```{r}
small_superheros <- select(superhero_info, name, alignment, height,weight) %>% 
  filter(weight <50) %>% 
  arrange(weight) %>% 
  mutate(height_weight_ratio=height/weight)
small_superheros
```

Using the mutate function, we can add a new column to our dataframe using already existing data. We can recall that ratios are simply a division problem between 2 values. In our case, we wish to find the height to weight ratio, as such, we would want to divide height by weight to obtain this ratio in a new column (to our pre-existing dataframe) termed `height_weight_ratio`.

12. Who has the highest height to weight ratio?  
```{r}
small_superheros <- select(superhero_info, name, alignment, height,weight) %>% 
  filter(weight <50) %>% 
  arrange(weight) %>% 
  mutate(height_weight_ratio=height/weight) %>% 
  arrange(desc(height_weight_ratio))
small_superheros
```
As shown, Groot (a personal favorite of mine) has the largest height to weight ratio. 

## `superhero_powers`
Have a quick look at the `superhero_powers` data frame.  
```{r}
glimpse(superhero_powers)
```

13. How many superheros have a combination of agility, stealth, super_strength, stamina?
```{r}
superhero_powers <- clean_names(superhero_powers)
```

```{r}
fancy_superheros <- select(superhero_powers, hero_names,agility,stealth,super_strength,stamina) %>% 
  filter(agility==TRUE) %>% 
  filter(stealth==TRUE) %>% 
  filter(super_strength==TRUE) %>% 
  filter(stamina==TRUE)
fancy_superheros
```
```{r}
count(fancy_superheros)
```
Under the specifications, we have 40 superheros with those powers. 

## Your Favorite
14. Pick your favorite superhero and let's see their powers!  
```{r}
superhero_powers %>% 
  filter(hero_names=="Iron Man") %>% 
  select_if(all)
```
The superpowers Iron Man possesses are shown above.  

15. Can you find your hero in the superhero_info data? Show their info! 
```{r}
superhero_info %>% 
  filter(name=="Iron Man")
```
All data about Iron Man in the superhero_info is shown above.

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.  