---
title: "summarize practice, `count()`, `across()`"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    toc: yes
    toc_float: yes
    keep_md: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Produce clear, concise summaries using a variety of functions in `dplyr` and `janitor.`  
2. Use the `across()` operator to produce summaries across multiple variables.  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("skimr")
library("palmerpenguins")
```

## Review
The summarize() and group_by() functions are powerful tools that we can use to produce clean summaries of data. Especially when used together, we can quickly group variables of interest and save time. Let's do some practice with the [palmerpenguins(https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data to refresh our memory.
```{r}
glimpse(penguins)
```

As biologists, a good question that we may ask is how do the measured variables differ by island (on average)?
```{r}
names(penguins)
```
```{r}
levels(penguins$island)
```
Because islands is a factor we can use the levels command

```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(mean_island=mean(island))
```

```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(mean_body_mass=mean(body_mass_g),
            n=n())
```

Why do we have NA here? Do all of these penguins lack data?
```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(number_NA=sum(is.na(body_mass_g)))
```

Nope, we have just 1 penguin messing up the entire calculation 
Well, that won't work so let's remove the NAs and recalculate.
```{r}
penguins %>% 
  filter(!is.na(body_mass_g)) %>% #take out all the NA and only pick out the things with values. AKA pull out all the observations with a number and leave the NAs behind 
  group_by(island) %>% 
 summarize(mean_body_mass=mean(body_mass_g),
            n=n())
```

What if we are interested in the number of observations (penguins) by species and island?
```{r}
penguins %>% 
  group_by(species, island) %>% 
  summarize(n=n(), .groups= 'keep')#the .groups argument here just prevents a warning message
```
`n=n()` pulls out the total number of individuals
Group by is only going to work with a categorical variable, we have to give it categories 

continuous variable is a number 


## Counts
Although these summary functions are super helpful, oftentimes we are mostly interested in counts. The [janitor package](https://garthtarr.github.io/meatR/janitor.html) does a lot with counts, but there are also functions that are part of dplyr that are useful.  

AKA how many ___. 

`count()` is an easy way of determining how many observations you have within a column. It acts like a combination of `group_by()` and `n()`.
```{r}
penguins %>% 
  count(island, sort = T) #sort=T sorts the column in descending order
```
Here we want to find out the number of penguins in each of the islands 

```{r}
tabyl(penguins$island)
```


Compare this with `summarize()` and `group_by()`.
```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(n=n())
```

You can also use `count()` across multiple variables.
```{r}
penguins %>% 
  count(island, species, sort = T) # sort=T will arrange in descending order
```
You can count the number of individuals in each species catgory in each island catgory. You can have multiple variables in the count. 

For counts, I also like `tabyl()`. Lots of options are supported in (Tabyl is part of janitor)  [tabyl](https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html)
```{r}
penguins %>% 
  tabyl(island,species)
```
You can also specify different variables with tabyl as well. 

## Practice
1. How does the mean of `bill_length_mm` compare between penguin species?
```{r}
penguins%>% 
  group_by(species, bill_length_mm) %>% 
  summarize(mean_bill_length= mean(bill_length_mm,na.RM=T)) ##notice that the na.Rm=T is inside the command taking the mean because this is specifically where we want to remove the NA from for the calculations we want to do 
```

2. For some penguins, their sex is listed as NA. Where do these penguins occur?
```{r}
penguins %>% 
  count(island,sex)
```
```{r}
penguins %>% 
  group_by(island,sex)
```
Notice the difference. Group by gives you everything. And count gives you just the specified vairbales and their total counts

## `across()`
What about using `filter()` and `select()` across multiple variables? There is a function in dplyr called `across()` which is designed to work across multiple variables. Have a look at Rebecca Barter's awesome blog [here](http://www.rebeccabarter.com/blog/2020-07-09-across/).    

What if we wanted to apply `summarize()` in order to produce distinct counts over multiple variables; i.e. species, island, and sex? Although this isn't a lot of coding you can image that with a lot of variables it would be cumbersome.
```{r}
penguins %>%
  summarize(distinct_species = n_distinct(species),
            distinct_island = n_distinct(island),
            distinct_sex = n_distinct(sex))
```

By using `across()` we can reduce the clutter and make things cleaner. Instead of using summarize seperatly for the variables we want 
```{r}
penguins %>%
  summarize(across(c(species, island, sex), n_distinct))
```
n_distinct is used to count the number of distinct observations for these variables 
This is very helpful for continuous variables.
```{r}
penguins %>%
  summarize(across(contains("mm"), mean, na.rm=T))
```
Looking for all the vairables with the "mm" in it 

```{r}
penguins %>%
  summarize(across(contains("mm"), \(x) mean(x, na.rm = TRUE))) ##Use this to correct the error R may give (the \(x)) thingy 
```

`group_by` also works.
```{r}
penguins %>%
  group_by(sex) %>% 
  summarize(across(contains("mm"), mean, na.rm=T))
```

Here we summarize across all variables.
```{r}
penguins %>%
  summarise_all(mean, na.rm=T)
```

Operators can also work, here I am summarizing across all variables except `species`, `island`, `sex`, and `year`.
```{r}
penguins %>%
  summarise(across(!c(species, island, sex, year), 
                   mean, na.rm=T))
```

All variables that include "bill"...all of the other dplyr operators also work.
```{r}
penguins %>%
  summarise(across(starts_with("bill"), mean, na.rm=T))
```

## Practice
1. Produce separate summaries of the mean and standard deviation for bill_length_mm, bill_depth_mm, and flipper_length_mm for each penguin species. Be sure to provide the number of samples.  
```{r}

```

## Wrap-up  

Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)