---
title: "`summarize()` and `group_by()`"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    toc: yes
    toc_float: yes
    keep_md: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Use a combination of `select()`, `filter()`, and `mutate()` to transform data frames.  
2. Use the `skimr` package to produce summaries of data.  
3. Produce clean summaries of data using `summarize()`.  
4. Use `group_by()` in combination with `summarize()` to produce grouped summaries of data.  

## Load the tidyverse and janitor
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Install `skimr`
```{r}
#install.packages("skimr")
library("skimr")
```

## Load the data
For this lab, we will use the built-in data on mammal sleep patterns. From: _V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007_.
```{r}
?msleep #putting a`?` in front of embedded datasets tells you more information about it 
names(msleep)
msleep <- msleep
```

We will also use [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data in the second part of lab.
```{r}
library("palmerpenguins")
```

## dplyr Practice
Let's do a bit more practice to make sure that we understand `select()`, `filter()`, and `mutate()`. Start by building a new data frame `msleep24` from the `msleep` data that: contains the `name` and `vore` variables along with a new column called `sleep_total_24` which is the amount of time a species sleeps expressed as a proportion of a 24-hour day. Restrict the `sleep_total_24` values to less than or equal to 0.3. Arrange the output in descending order.  
```{r}
summary(msleep)
```

```{r}
msleep24<- msleep %>% 
  mutate(sleep_total_24= sleep_total/24) %>% 
  select(name,vore, sleep_total, sleep_total_24) %>%
  filter(sleep_total_24<=0.3) %>% 
  arrange(desc(sleep_total_24))

msleep24
```


Did `dplyr` do what we expected? How do we check our output? Remember, just because your code runs it doesn't mean that it did what you intended.
```{r}
summary(msleep24)
```
First check that the minimum and maximium values align with your requirment 
Try out the new function `skim()` as part of the `skimr` package.
```{r}
skim(msleep24)
```
Skim is another way to get an overview of the data, similar to summary and glimpse 
Histograms are also a quick way to check the output.
```{r}
hist(msleep24$sleep_total_24)
```
Highest value being 0.30 is good since it is was we specified in the code. In this `hist` command you have to specify the specific variable you want to make a histogram of. 


## `summarize()`
`summarize()` will produce summary statistics for a given variable in a data frame. For example, if you are asked to calculate the mean of `sleep_total` for large and small mammals you could do this using a combination of commands, but it isn't very efficient or clean. We can do better!  
```{r}
head(msleep)
```

For example, if we define "large" as having a `bodywt` greater than 200 then we get the following:
```{r}
large <- msleep %>% 
  select(name, genus, bodywt, sleep_total) %>% 
  filter(bodywt > 200) %>% 
  arrange(desc(bodywt))
large
```

Rm is a way to remove something from the enviornment panel in case you overrode the code somewhere else
```{r}
mean(large$sleep_total)
```

We can accomplish the same task using the `summarize()` function to make things cleaner.
```{r}
msleep %>% 
  filter(bodywt>200) %>% 
  summarize(mean_sleep_lg=mean(sleep_total))
```

You can also combine functions to make useful summaries for multiple variables.
```{r}
msleep %>% 
  filter(bodywt>200) %>% 
  summarize(mean_sleep_lg=mean(sleep_total),
            min_sleep_lg=min(sleep_total),
            max_sleep_lg=max(sleep_total),
            sd_sleep_lg=sd(sleep_total), 
            total=n()) ##This tells you the total number of observations, aka the number of rows 
            
```

## Practice
1. What is the mean, min, and max `bodywt` for the taxonomic order Primates? Provide the total number of observations.

```{r}
names(msleep)
```

```{r}
msleep %>% 
  filter(order=="Primates") %>% 
  summarize(mean_bodywt=mean(bodywt),
            min_bodywt=min(bodywt),
            max_bodywt=max(bodywt),
            total=n()) ##This tells you the total number of observations, aka the number of rows 
            

```

`n_distinct()` is a very handy way of cleanly presenting the number of distinct observations. Here we show the number of distinct genera over 100 in body weight.  

Notice that there are multiple genera with over 100 in body weight. 
```{r}
msleep %>% 
  filter(bodywt > 100)
```

n_distinct() is a very handy way of cleanly presenting the number of distinct observations. Here we show the number of distinct genera over 100 in body weight.THIS IS ON THE EXAM 

Way to pull out unique values in a category of data. Difficult to get a count of the column since there are duplicates. 
```{r}
msleep %>% 
  summarize(n_genera=n_distinct(genus)) ##this is going to count the number of genera in msleep 
```
```{r}
glimpse(msleep)
```
Every row should be its own unique data Because distinct gave back less than the total number of row wehave confirmation that there were indeed repeats. 


There are many other useful summary statistics, depending on your needs: sd(), min(), max(), median(), sum(), n() (returns the length of a column), first() (returns first value in a column), last() (returns last value in a column) and n_distinct() (number of distinct values in a column).

## Practice
1. How many genera are represented in the msleep data frame?
```{r}
msleep %>% 
  tabyl(genus)
```
Tabyl is another way to get the number of observations, and check if there are any duplicates 
```{r}

```

2. What are the min, max, and mean `sleep_total` for all of the mammals? Be sure to include the total n.
```{r}
msleep %>% 
  summarize(mean_sleep_all=mean(sleep_total),
            min_sleep_all=min(sleep_total),
            max_sleep_all=max(sleep_total),
            total=n())
```

## `group_by()`
The `summarize()` function is most useful when used in conjunction with `group_by()`. Although producing a summary of body weight for all of the mammals in the data set is helpful, what if we were interested in body weight by feeding ecology?

Group together categorical variables before doing the summary command 
```{r}
msleep %>%
  group_by(vore) %>% #we are grouping by feeding ecology, a categorical variable. Ie what do these specific values mean for omni, herb, and carn (by specifing the vore variable/column it gives you the min,mean,max for each vore type)
  summarize(min_bodywt = min(bodywt),
            max_bodywt = max(bodywt),
            mean_bodywt = mean(bodywt),
            total=n())
```
```{r}
names(msleep)
```

## Practice
1. Calculate mean brain weight by taxonomic order in the msleep data.
```{r}
msleep %>% 
  group_by(order) %>%
  summarize(min_brainwt = min(brainwt),
            max_brainwt = max(brainwt),
            mean_brainwt = mean(brainwt),
            total=n()) %>% 
  arrange(desc(mean_brainwt))
```

2. What does `NA` mean? How are NA's being treated by the summarize function?
```{r}
msleep %>% 
  filter(order=="Carnivora") %>% 
  select(order, genus,brainwt)

```

3. Try running the code again, but this time add `na.rm=TRUE`. What is the problem with Cetacea? Compare this to Carnivora. 
```{r}
msleep %>% 
  group_by(order) %>% 
  summarise(mean_brain_wt=mean(brainwt,na.rm = T))
```
```{r}
msleep %>% 
  filter(order=="cetacea") %>% 
  select(order, genus,brainwt)
```


## That's it! Let's take a break and then move on to part 2! 

-->[Home](https://jmledford3115.github.io/datascibiol/)  